---
title: "Differential Virulence Report for `r snakemake@params$exp` (`r snakemake@params$dir`)"
author: "[iPoolSeq-Pipeline](http://www.cibiv.at/software/ipoolseq-pipeline) `r snakemake@params$version`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmdformats::readthedown:
    highlight: kate
# ipoolseq.differential.virulence.Rmd, Copyright 2016, 2017, 2018, 2019 Florian G. Pflug
# 
# This file is part of the iPool-Seq Analysis Pipeline
#
# The iPool-Seq Analysis Pipeline is free software: you can redistribute it
# and/or modify it under the terms of the GNU Affero General Public License as
# published by the Free Software Foundation, either version 3 of the License,
# or (at your option) any later version.
#
# The iPool-Seq Analysis Pipeline is distributed in the hope that it will be
# useful, but WITHOUT ANY WARRANTY; without even the implied warranty of
# MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
# GNU Affero General Public License for more details.
#
# You should have received a copy of the GNU Affero General Public License
# along with the iPool-Seq Analysis Pipeline.  If not, see
# <http://www.gnu.org/licenses/
---

<style>
/* With readthedown, the author name is printed on a dark background,
 * making the default dark red link color illegible. Use white instead,
 * and use the same hover effect as in the main document.
 */
#postamble .author a {
  color: white;
  background-image: linear-gradient(180deg,#d64a70,#d64a70);
  background-size: 100% 100%;
  background-repeat: no-repeat;
  background-position: 0 2em;
  font-weight: 300;
  padding: .125em 1px 0 1px;
  transition: background-position .15s, color .15s
}
#postamble .author a:hover {
  text-decoration: none;
  background-position: 0 0.3em;
}
</style>

```{r stub, include=FALSE, eval=FALSE}
# Set "eval=TRUE" above to run this chunk interactively in R Studio,
# and adjust the input and output files to point to the sample you
# want to test with
design <- "Transposon"
replicate <- "r77647"
setClass("SnakemakeStub", representation(input = "list", params = "list", output = "list"))
snakemake <- new("SnakemakeStub",
                 input=list(ann=paste("..", "cfg", design, "annotation.gff3.gz", sep="/"),
                            isites_in=paste("..", "data", design, paste0(replicate, "-in.isites.gff3.gz"), sep="/"),
                            isites_out=paste("..", "data", design, paste0(replicate, "-out.isites.gff3.gz"), sep="/"),
                            pool_in_5p=paste("..", "data", design, paste0(replicate, "-in:5p.count.tab"), sep="/"),
                            pool_in_3p=paste("..", "data", design, paste0(replicate, "-in:3p.count.tab"), sep="/"),
                            pool_out_5p=paste("..", "data", design, paste0(replicate, "-out:5p.count.tab"), sep="/"),
                            pool_out_3p=paste("..", "data", design, paste0(replicate, "-out:3p.count.tab"), sep="/"),
                            stats_in_5p=paste("..", "data", design, paste0(replicate, "-in:5p.stats.tab"), sep="/"),
                            stats_in_3p=paste("..", "data", design, paste0(replicate, "-in:3p.stats.tab"), sep="/"),
                            stats_out_5p=paste("..", "data", design, paste0(replicate, "-out:5p.stats.tab"), sep="/"),
                            stats_out_3p=paste("..", "data", design, paste0(replicate, "-out:3p.stats.tab"), sep="/"),
                            trumicount_pdf_in_5p=paste("..", "data", design, paste0(replicate, "-in:5p.count.pdf"), sep="/"),
                            trumicount_pdf_in_3p=paste("..", "data", design, paste0(replicate, "-in:3p.count.pdf"), sep="/"),
                            trumicount_pdf_out_5p=paste("..", "data", design, paste0(replicate, "-out:5p.count.pdf"), sep="/"),
                            trumicount_pdf_out_3p=paste("..", "data", design, paste0(replicate, "-out:3p.count.pdf"), sep="/"),
                            fastqc_html_in_5p_r1=paste("..", "data", design, paste0(replicate, "-in:5p.fastqc.1.html"), sep="/"),
                            fastqc_html_in_5p_r2=paste("..", "data", design, paste0(replicate, "-in:5p.fastqc.2.html"), sep="/"),
                            fastqc_html_in_3p_r1=paste("..", "data", design, paste0(replicate, "-in:3p.fastqc.1.html"), sep="/"),
                            fastqc_html_in_3p_r2=paste("..", "data", design, paste0(replicate, "-in:3p.fastqc.2.html"), sep="/"),
                            fastqc_html_out_5p_r1=paste("..", "data", design, paste0(replicate, "-out:5p.fastqc.1.html"), sep="/"),
                            fastqc_html_out_5p_r2=paste("..", "data", design, paste0(replicate, "-out:5p.fastqc.2.html"), sep="/"),
                            fastqc_html_out_3p_r1=paste("..", "data", design, paste0(replicate, "-out:3p.fastqc.1.html"), sep="/"),
                            fastqc_html_out_3p_r2=paste("..", "data", design, paste0(replicate, "-out:3p.fastqc.2.html"), sep="/")),
                 params=list(version={f <- file("../VERSION", "r"); v <- readLines(f, n=1); close(f); v},
                             dir=design, exp=replicate),
                 output=list(table=paste("..", "data", design, paste0(replicate, "dv.tab"), sep="/")))
```

```{r setup, include=FALSE}
# Load output libraries
library(knitr)
library(rmdformats)

# Configure knitr
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	echo = FALSE,
	cache = FALSE
)

# Load libraries
library(data.table)
library(DT)
library(rtracklayer)
library(plotly)
source("scripts/ipoolseq.model.R")

# Output '-' for NA values
options(knitr.kable.NA = '-')

# Set FDR to 5%
PLOT.FDR.THRESHOLD <- 0.05

# Convert a p-value into a significance marker
#   '?'     if p notin [0, 1]
#   ' '     if p in (1e-1, 1]
#   '+'     if p in (5e-2, 1e-1]
#   '*'     if p in (1e-2, 5e-1]
#   '**'    if p in (1e-3, 1e-2]
#   '***'   if p in (1e-4, 1e-3]
#   '****'  if p in (1e-5, 1e-4]
#   '*****' if p in [0,    1e-5]
sig.stars <- function(p) {
  pp <- ifelse((p > 0) & (p < 1), p, 0.5)
  ifelse(!is.finite(p) | (p < 0) | (p > 1), "?",
         ifelse(p > 0.1, " ",
                ifelse(p > 0.05, "+",
                       ifelse(p == 0, "*****",
                              strrep("*", pmin(ceiling(-log10(pp)*(1+.Machine$double.eps))-1, 5))))))
}

# Create a Markdown link
make_link <- function(path, text) {
  paste0('[', if(missing(text)) path else text, '](', URLencode(path, reserved=TRUE), ')')
}

# Correlation with zero intercept
cor.intercept0 <- function(x, y) {
  i <- !is.na(x) & !is.na(y)
  k <- sum(x[i] * y[i]) / sum(x[i]**2)
  r.sq <- 1 - sum((y[i] - k*x[i])**2) / sum((y[i] - mean(y[i]))**2)
  sqrt(r.sq)
}
```

```{r load}
# Loading gene annotation
message("Loading gene annotations from ", snakemake@input$ann)
annotations <- readGFFAsGRanges(snakemake@input$ann, version=3)
genes <- annotations[annotations$type=="gene"]

# Load list of knockouts, find assumed neutral set
message("Loading insertion sites from ", snakemake@input$isites_in)
knockouts <- readGFFAsGRanges(snakemake@input$isites_in, version=3)
# Assign knockouts to genes
local({
  hits <- findOverlaps(knockouts, genes, ignore.strand=TRUE)
  knockouts$GeneIndex <<- NA_integer_
  knockouts$GeneIndex[queryHits(hits)] <<- subjectHits(hits)
})
# Mark knockouts not assigned to genes as neutral
knockouts$Neutral <- ifelse(is.na(knockouts$GeneIndex), 1, 0)

# Find neutral knockouts!
NEUTRAL <- knockouts$Name[knockouts$Neutral == 1]

# Load UMI count tables produced by TRUmiCount
# Note that TRUmiCount thinks in terms of RNA-Seq and calls whatever key it used
# to group and count reads a "gene". In our case that key is the cassette insertion
# site, which is therefore stored in the column "gene" in TRUmiCount's output.
load.trumicount.output <- function(file, flank) {
  message("Loading TRUmiCount results from ", file)
  t <- data.table(read.table(file, header=TRUE, sep='\t'))
  # Rename column "gene" to "knockout"
  colnames(t) <- ifelse(colnames(t) == "gene", "knockout", colnames(t))
  # Make sure there is exactly one row per knockout and add flank
  allKOs <- data.table(knockout=knockouts$Name)
  t <- t[allKOs,, on="knockout"]
  t[, flank := flank ]
}
counts.flank.in.5p <- load.trumicount.output(snakemake@input[["pool_in_5p"]], "5p")
counts.flank.in.3p <- load.trumicount.output(snakemake@input[["pool_in_3p"]], "3p")
counts.flank.out.5p <- load.trumicount.output(snakemake@input[["pool_out_5p"]], "5p")
counts.flank.out.3p <- load.trumicount.output(snakemake@input[["pool_out_3p"]], "3p")

# Combine 5' and 3' data into a single table
counts.flank.in <- rbind(counts.flank.in.5p, counts.flank.in.3p)
counts.flank.out <- rbind(counts.flank.out.5p, counts.flank.out.3p)

join.flanks.as.columns <- function(data) {
  data[, list(n5p=n.obs[flank=="5p"],
              l5p=loss[flank=="5p"],
              n3p=n.obs[flank=="3p"],
              l3p=loss[flank=="3p"])
       , by="knockout"]
}

# Aggregate the data for the 5' and 3' flank of each gene.
# We add the counts and average the losses of the two flanks for each gene,
aggregate.flanks <- function(counts.flank) {
  counts.flank[, list(
    n = sum(n.obs, na.rm=TRUE),
    loss = mean(loss, na.rm=TRUE)
  ), by="knockout"]
}
counts.in <- aggregate.flanks(counts.flank.in)
counts.out <- aggregate.flanks(counts.flank.out)

# Combine input and output pool data into a single table
counts <- merge(counts.in, counts.out, by="knockout", suffixes=c(".in", ".out"))[,list(
  knockout,
  is.neutral=knockout %in% NEUTRAL,
  gene=NA_integer_,
  n.in, loss.in,
  n.out, loss.out,
  abundance.in=ifelse(n.in > 0, n.in / (2 * (1-loss.in)), 0),
  abundance.out=ifelse(n.out > 0, n.out / (2 * (1-loss.out)), 0)
)]

# Fill in index of the gene (if any) affected by a knockout
local({
  allKOmeta <- data.table(knockout=knockouts$Name, gene=knockouts$GeneIndex)
  counts[, "gene" := allKOmeta[counts, gene, on=c("knockout")] ]
})
```

# 1. Files

The following table lists the main input files used to compute differential virulences, and the name of the generated results table. The results table lists, for each knockout, the change in virulence (as a log2 fold change of normalized output abundance relative to the neutral reference set), p-values, and FDR-corrected (using the method of Benjamini-Hochberg) q-values for the significance of the change.

```{r table_files}
knitr::kable(data.frame(row.names=c("list of insertional knockouts (mutants)",
                                    "5' knockout abundances in (pre-infection)  input pool",
                                    "3' knockout abundances in (pre-infection)  input pool",
                                    "5' knockout abundances in (post-infection) output pool",
                                    "3' knockout abundances in (post-infection) output pool",
                                    "differential virulence results table"),
                        file=c(snakemake@input$isites_in,
                               make_link(basename(snakemake@input$pool_in_5p)),
                               make_link(basename(snakemake@input$pool_in_3p)),
                               make_link(basename(snakemake@input$pool_out_5p)),
                               make_link(basename(snakemake@input$pool_out_3p)),
                               make_link(basename(snakemake@output$table))),
                        report=c("",
                                 make_link(basename(snakemake@input$trumicount_pdf_in_5p), "TRUmiCount report"),
                                 make_link(basename(snakemake@input$trumicount_pdf_in_3p), "TRUmiCount report"),
                                 make_link(basename(snakemake@input$trumicount_pdf_out_5p), "TRUmiCount report"),
                                 make_link(basename(snakemake@input$trumicount_pdf_out_3p), "TRUmiCount report"),
                                 "")))
```

# 2. Quality Control

## Sequencing and Trimming

[FastQC](http://www.bioinformatics.babraham.ac.uk/projects/fastqc) reports for the trimmed reads. There should be little or no remaining adapter content, and per-base qualities should not drop too much at the end of the reads, otherwise further trimming might be required.

```{r table_fastqc}
knitr::kable(data.frame(
  row.names=c("5' Read 1", "5' Read 2", "3' Read 1", "3' Read 2"),
  `Input pool` =c(make_link(basename(snakemake@input$fastqc_html_in_5p_r1), "FastQC report"),
                  make_link(basename(snakemake@input$fastqc_html_in_5p_r2), "FastQC report"),
                  make_link(basename(snakemake@input$fastqc_html_in_3p_r1), "FastQC report"),
                  make_link(basename(snakemake@input$fastqc_html_in_3p_r2), "FastQC report")),
  `Output pool`=c(make_link(basename(snakemake@input$fastqc_html_out_5p_r1), "FastQC report"),
                  make_link(basename(snakemake@input$fastqc_html_out_5p_r2), "FastQC report"),
                  make_link(basename(snakemake@input$fastqc_html_out_3p_r1), "FastQC report"),
                  make_link(basename(snakemake@input$fastqc_html_out_3p_r2), "FastQC report")),
  check.names=FALSE
))
```

## Read and UMI count statistics

The number of remaining read pairs and the number of unique UMIs within these pairs after each analysis step. The percentages are relative to the *first* number within each column. Since unique UMIs are determined for the 5' and 3' flank of each knockout, no UMI counts are reported until reads have been assigned to the individual knockouts (and their flanks).

Also check the *TRUmiCount report* (found above under *Files*) for details about the performance of the [TRUmiCount](http://www.cibiv.at/software/trumicount) UMI filtering step.

```{r table_stats_in}
message("Loading statistics from ", snakemake@input$stats_in_5p)
stats_in_5p <- read.table(snakemake@input$stats_in_5p, header=TRUE, sep="\t", comment.char='', na.strings='-', check.names=FALSE)
message("Loading statistics from ", snakemake@input$stats_in_3p)
stats_in_3p <- read.table(snakemake@input$stats_in_3p, header=TRUE, sep="\t", comment.char='', na.strings='-', check.names=FALSE)

stopifnot(all(stats_in_5p$`After Step` == stats_in_3p$`After Step`))
stats <- data.frame(row.names=stats_in_5p$`After Step`,
                    `#5\' Reads Input`=stats_in_5p$`#Reads`,
                    `#3\' Reads Input`=stats_in_3p$`#Reads`,
                    `#5\' UMIs Input`=stats_in_5p$`#UMIs`,
                    `#3\' UMIs Input`=stats_in_3p$`#UMIs`,
                    check.names=FALSE)
for(c in c('#5\' Reads Input', '#3\' Reads Input', '#5\' UMIs Input', '#3\' UMIs Input')) {
  stats[[c]] <- ifelse(!is.na(stats[[c]]),
                       paste0(stats[[c]], " (",
                              round(100*stats[[c]]/(stats[[c]][min(which(!is.na(stats[[c]])))])), "%)"),
                       NA)
}
knitr::kable(stats)
```

```{r table_stats_out}
message("Loading statistics from ", snakemake@input$stats_out_5p)
stats_out_5p <- read.table(snakemake@input$stats_out_5p, header=TRUE, sep="\t", comment.char='', na.strings='-', check.names=FALSE)
message("Loading statistics from ", snakemake@input$stats_out_3p)
stats_out_3p <- read.table(snakemake@input$stats_out_3p, header=TRUE, sep="\t", comment.char='', na.strings='-', check.names=FALSE)

stopifnot(all(stats_out_5p$`After Step` == stats_out_3p$`After Step`))
stats <- data.frame(row.names=stats_out_5p$`After Step`,
                    `#5\' Reads Output`=stats_out_5p$`#Reads`,
                    `#3\' Reads Output`=stats_out_3p$`#Reads`,
                    `#5\' UMIs Output`=stats_out_5p$`#UMIs`,
                    `#3\' UMIs Output`=stats_out_3p$`#UMIs`,
                    check.names=FALSE)
for(c in c('#5\' Reads Output', '#3\' Reads Output', '#5\' UMIs Output', '#3\' UMIs Output')) {
  stats[[c]] <- ifelse(!is.na(stats[[c]]),
                       paste0(stats[[c]], " (",
                              round(100*stats[[c]]/(stats[[c]][min(which(!is.na(stats[[c]])))])), "%)"),
                       NA)
}
knitr::kable(stats)
```

## Correlation of 3' and 5' Flank Abundances

Since the abundance of each knockout is measured twice, once for the 5' and once for the 3' flank of the knockout cassette integration, the correlation of these two measurements provides a quality check of the data. The following table lists the correlation of both the raw, and the TRUmiCount-corrected UMI counts detected for the 5' and 3' flanks of each knockout. Correlations are computed from the *square roots* of counts respectively abundances (to avoid overdue influence of large counts), and with the intercept forced to zero (regression line passing through the origin).

```{r table_corr_flanks}
knitr::kable(data.table(
  `Pool`=c(
  "input",
  "output",
  "input",
  "output"
  ),
  Type=c(
  "Raw (after UMI-Tools and read-count threshold)",
  "Raw (after UMI-Tools and read-count threshold)",
  "Loss-corrected (after TRUmiCount)",
  "Loss-corrected (after TRUmiCount)"
  ),
  `Correlation 5' vs. 3 (intercept zero)'`=c(
  join.flanks.as.columns(counts.flank.in)[, cor.intercept0(sqrt(n5p), sqrt(n3p))],
  join.flanks.as.columns(counts.flank.out)[, cor.intercept0(sqrt(n5p), sqrt(n3p))],
  join.flanks.as.columns(counts.flank.in)[, cor.intercept0(sqrt(n5p/(1-l5p)), sqrt(n3p/(1-l3p)))],
  join.flanks.as.columns(counts.flank.out)[, cor.intercept0(sqrt(n5p/(1-l5p)), sqrt(n3p/(1-l3p)))])
), digits=3)
```

The following plots show the correlation of 5' and 3' UMI counts (again raw and loss-corrected by TRUmiCount) in more detail.

```{r plot_corr_flanks, fig.height=6, fig.width=8}
p <- function(n.obs.5p, n.obs.3p, labels, type, first=FALSE, main) {
  missing <- !is.finite(n.obs.5p) | (n.obs.5p == 0) | !is.finite(n.obs.3p) | (n.obs.3p == 0)
  plot_ly(showlegend=first) %>%
  add_markers(x=n.obs.5p[!missing],
              y=n.obs.3p[!missing],
              color=I("black"),
              text=labels[!missing], name="UMIs found for 5' and 3' flank", legendgroup="nz",
              marker=list(symbol="circle-open")) %>%
  add_markers(x=pmax(n.obs.5p[missing], 0.5, na.rm=TRUE),
              y=pmax(n.obs.3p[missing], 0.5, na.rm=TRUE),
              color=I("black"),
              text=labels[missing], name="No UMIs found for 5' or 3' flank", legendgroup="z",
              marker=list(symbol="x")) %>%
  layout(xaxis=list(type="log", title=paste0(type, " #UMIs 5'")),
         yaxis=list(type="log", title=paste0(type, " #UMIs 3'")),
         annotations=list(x=0.4, y=1, text = main, xanchor="center", yanchor="bottom",
                          xref='paper', yref='paper',
                          showarrow=F, font=list(size=15)))
}

p.raw.in <- join.flanks.as.columns(counts.flank.in)[, {
  p(n5p, n3p, knockout, main="Input 5' vs. 3' flank (raw)", type="Raw", first=TRUE)
}]
p.raw.out <- join.flanks.as.columns(counts.flank.out)[, {
  p(n5p, n3p, knockout, main="Output 5' vs. 3' flank (raw)", type="Raw")
}]
p.corr.in <- d <- join.flanks.as.columns(counts.flank.in)[, {
  p(n5p/(1-l5p), n3p/(1-l3p), knockout, main="Input 5' vs. 3' flank (loss-corr.)", type="Loss-corr.")
}]
p.corr.out <- d <- join.flanks.as.columns(counts.flank.out)[, {
  p(n5p/(1-l5p), n3p/(1-l3p), knockout, main="Output 5' vs. 3' flank (loss-corr.)", type="Loss-corr.")
}]
subplot(list(p.raw.in, p.raw.out, p.corr.in, p.corr.out), nrows=2,
        titleX=TRUE, titleY=TRUE, margin=0.12) %>%
layout(width=600, height=600,
       margin=list(l=0, r=0, t=50, b=50),
       legend=list(orientation="h", bgcolor="lightgrey",
                   x=0, y=-0.2, xanchor="left", yanchor="top"))
```

__________
(Zero detected UMIs are shown in these logarithmic plots with a cross "x" instead of an open circle "o")

## Correlation of Input and Output Abundances

In iPool-Seq-based screens, the abundances of the different knockout strains are often spread over multiple orders of magnitude, and differences in input abundances can thus affect the abundance in the output more strongly than the mutant's phenotypes. The statistical model used to detect significant changes of virulence must thus take the input abundance into account, and assumes that for neutral knockouts, input and output abundances are proportional.

The following table shows the observed correlation of (loss-corrected) input and output abundances (averaged across the 5' and 3' flank measurements). Correlations are computed from the *square roots* of counts respectively abundances (to avoid overdue influence of large counts), and with the intercept forced to zero (regression line passing through the origin).

```{r table_corr_io}
knitr::kable(data.table(
  Type=c(
  "Raw (after UMI-Tools and read-count threshold)",
  "Loss-corrected (after TRUmiCount)"
  ),
  `Correlation Input vs. Output`=c(
  counts[, cor.intercept0(sqrt(n.in), sqrt(n.out))],
  counts[, cor.intercept0(sqrt(abundance.in), sqrt(abundance.out))])
), digits=3)
```

The following plots show the correlation of input and output abundances in more detail.

```{r plot_corr_io, fig.height=3, fig.width=8}
p <- function(abd.in, abd.out, labels, type, first=FALSE, main) {
  missing <- !is.finite(abd.in) | (abd.in == 0) | !is.finite(abd.out) | (abd.out == 0)
  o <- plot_ly(showlegend=first)
  o <- o %>% add_markers(x=abd.in[!missing],
                         y=abd.out[!missing],
                         color=I("black"),
                         text=labels[!missing], name="present in input and output", legendgroup="nz",
                         marker=list(symbol="circle-open"))
  if (any(missing))
    o <- o %>% add_markers(x=pmax(abd.in[missing], 0.5, na.rm=TRUE),
                           y=pmax(abd.out[missing], 0.5, na.rm=TRUE),
                           color=I("black"),
                           text=labels[missing], name="missing in input or output", legendgroup="z",
                          marker=list(symbol="x"))
  o <- o %>% layout(xaxis=list(type="log", title=paste0(type, " #UMIs in input")),
                    yaxis=list(type="log", title=paste0(type, " #UMIs in output")),
                    annotations=list(x=0.4, y=1, text=main, xanchor="center", yanchor="bottom",
                                    xref='paper', yref='paper',
                                    showarrow=F, font=list(size=15)))
  o
}

p.raw <- counts[, {
  p(n.in, n.out, knockout, main="Input vs. Output (raw)", type="Raw", first=TRUE)
}]
p.corr <- counts[, {
  p(abundance.in, abundance.out, knockout, main="Input vs. Output (loss-corr.)", type="Loss-corr.")
}]
subplot(list(p.raw, p.corr), nrows=1,
        titleX=TRUE, titleY=TRUE, margin=0.12) %>%
layout(width=600, height=300,
       margin=list(l=0, r=0, t=50, b=50),
       legend=list(orientation="h", bgcolor="lightgrey",
                   x=0, y=-0.4, xanchor="left", yanchor="top"))
```

__________
(Zero detected UMIs are shown in these logarithmic plots with a cross "x" instead of an open circle "o")

# 3. Results

## Statistical Model and Parameters

The (raw) output UMI count $N_\text{out}$ of a neutral knockout given its (raw) input UMI count $N_\text{in}$ and TRUmiCount-estimated loss (i.e. fraction of lost UMIs) $\ell_\text{in}$ for the input and $\ell_\text{out}$ for the output is modelled with the following negative binomial model

$$
N_\text{out} \,|\, N_\text{in} \;\sim\; \text{NegBin}\left(\mu:=N_\text{in}\cdot\lambda\cdot\frac{1-\ell_\text{out}}{1-\ell_\text{in}}, r:=\frac{N_\text{in}}{1+d\cdot N_\text{in}}\right).
$$

(UMI counts of 5' and 3' flanks are added to obtain $N_\text{in}$ and $N_\text{out}$, losses for 5' and 3' flanks are averaged to obtain $\ell_\text{in}$ and $\ell_\text{out}$). Parameter $\lambda$ measures the relative size of the output library (i.e. loss-corrected total number of UMIs per neutral knockout) compared to the input. Parameter $d$ represents the biological contribution to the squared [coefficient of variation](http://en.wikipedia.org/wiki/Coefficient_of_variation) of the (raw) output UMI counts $N_\text{out}$. The total squared coefficient of variation $\text{CV}^2 = \sigma^2 / \mu^2$ of $N_\text{out}$ (where $\mu$ is the mean and $\sigma^2$ the variance of $N_\text{out}$) comprises three contributors,

$$
  \text{CV}^2 = \frac{1}{\mu} + \frac{1}{N_\text{in}} + d,
$$

where the first two are of technical nature and represent the variation due to (Poissonian, i.e. non-exhaustive) sampling of genomes in the input ($1/N_\text{in}$) and output ($1/\mu$) pools, and the third ($d$) represents the biological variation due to growth differences between neutral mutants.

The optimal (likelihood-maximazing) values for parameters $\lambda$ and $d$ for the reference set of neutral knockouts are:

```{r fit_model, warning=TRUE}
# Counts to fit model against
counts.paramest <- counts[is.neutral & (n.in > 0) & (n.out > 0) & (loss.in < 1) & (loss.out < 1)]
# Fallback scaling factor (lambda) -- simply the quotient of the loss-corr. means
scale.fallback <- counts.paramest[, mean(n.out/(1-loss.out), na.rm=TRUE) / mean(n.in/(1-loss.in), na.rm=TRUE)]
null.model <- if (nrow(counts.paramest) >= 2) {
  # Enough datapoints to fit model
  tryCatch(counts.paramest[, ipsmodel.fit(n.out=n.out, n.in=n.in, l.out=loss.out, l.in=loss.in)],
           error=function(e) {
             # Error during fitting, warn and use fallback
             warning(paste0("Model parameter estimation failed: ", conditionMessage(e)),
                     call.=FALSE);
             data.table(scale=scale.fallback, disp=NA_real_)
  })
} else {
  # To few data points, warn and use fallback
  warning(paste0("Model parameter estimation failed: ",
                 "too few mutants in the neutral reference were detected in both ",
                 "input and output (need 2, found ", nrow(counts.paramest), ")"),
          call.=FALSE)
  data.table(scale=scale.fallback, disp=NA_real_)
}
knitr::kable(signif(null.model, digits=3),
             col.names=c('Relative output library size (lambda)', 'Biological contribution to squared CV (d)'))
```

## Differential virulence compared to the neutral reference set

For every knockout strain detected in both input and output with (raw) UMI counts $N_\text{in}$ respectively $N_\text{out}$ (counts of 5' and 3' flanks added) and TRUmiCount-estimated  losses (i.e. fraction of lost UMIs) $\ell_\text{in}$ and $\ell_\text{out}$ (losses for 5' and 3' flanks averaged), the virulence log2 fold-change compared to the neutral reference set is

$$
\log_2 \Delta v = \log_2 \frac{N_\text{out} / (1-\ell_\text{out})}{\lambda\cdot N_\text{in} / (1-\ell_\text{in})}.
$$

($\lambda$ is the relative size of the outpout library compared to the input library, see *Statistical Model and Parameters*)

```{r sig_test}
# Compute log2 fold changes and p-values
results <- counts[, list(
  knockout, is.neutral,
  gene_id=ifelse(!is.na(gene), genes[pmax(gene, 1, na.rm=TRUE)]$ID, NA_character_),
  gene_name=ifelse(!is.na(gene), genes[pmax(gene, 1, na.rm=TRUE)]$Name, NA_character_),
  n.out, loss.out, abundance.out,
  n.in, loss.in, abundance.in,
  log2fc=ifelse(is.finite(abundance.out) & is.finite(abundance.in) & (abundance.in>0) & (abundance.out>0),
                log2(abundance.out / (null.model$scale * abundance.in)),
                NA_real_),
  low.pval=ifelse(is.finite(n.out) & is.finite(n.in) & (n.out > 0) & (n.in > 0),
                  pipsmodel(n.out, null.model$scale, null.model$disp, n.in, loss.out, loss.in),
                  NA_real_),
  high.pval=ifelse(is.finite(n.out) & is.finite(n.in) & (n.out > 0) & (n.in > 0),
                  1 - pipsmodel(n.out - 1, null.model$scale, null.model$disp, n.in, loss.out, loss.in),
                  NA_real_)
)]

# Compute q-values using the method of Benjamini & Hochberg (BH)
results[, low.qval := p.adjust(low.pval, method='BH')]
results[, high.qval := p.adjust(high.pval, method='BH')]

# Write output table
write.table(results, file=snakemake@output$table, col.names=TRUE, sep="\t", row.names=FALSE)
```

Due to the negative binomial model (see *Statistical Model and Parameters*), the power to detect changes in virulence increases with the input abundance of a knockout. The following plot shows the change in virulence (as log2 fold-change) against the input abundance, and indicates which knockout have a virulence significantly different from the neutral reference set at a *false discovery rate* (FDR) threshold of `r PLOT.FDR.THRESHOLD`.

```{r plot_log2fc}
# Compute (approximate) insigificance region
loss.out <- mean(results$loss.out, na.rm=TRUE)
loss.in <- mean(results$loss.in, na.rm=TRUE)
a.in <- 10**seq(from=0, to=2*log10(max(results$abundance.in, na.rm=TRUE)), length.out=200)
l2fc.low <- log2(pmax(qipsmodel(p=rep(PLOT.FDR.THRESHOLD, along.with=a.in),
                                scale=null.model$scale, disp=null.model$disp,
                                n.in=a.in * (1 - loss.in), l.out=loss.out, l.in=loss.in) - 1, 0) /
                 ((1 - loss.out) * a.in * null.model$scale))
l2fc.high <- log2(pmax(qipsmodel(p=rep(1 - PLOT.FDR.THRESHOLD, along.with=a.in),
                                 scale=null.model$scale, disp=null.model$disp,
                                 n.in=a.in * (1 - loss.in), l.out=loss.out, l.in=loss.in) + 1, 0) /
                 ((1 - loss.out) * a.in * null.model$scale))

plot_ly(width=600, height=600) %>%
add_trace(x=c(a.in, rev(a.in)),
          y=pmax(c(rev(cummin(rev(l2fc.low))),
                   cummax(rev(l2fc.high))),
                 10*min(results$log2fc, 0, na.rm=TRUE)),
          name='insignificant area (approximate, without FDR correction)',
          type='scatter', mode='none', fill='toself', fillcolor='lightgrey') %>%
add_trace(x=results[!is.neutral & (low.qval >= PLOT.FDR.THRESHOLD) & (high.qval >= PLOT.FDR.THRESHOLD), abundance.in],
          y=results[!is.neutral & (low.qval >= PLOT.FDR.THRESHOLD) & (high.qval >= PLOT.FDR.THRESHOLD), log2fc],
          text=results[!is.neutral & (low.qval >= PLOT.FDR.THRESHOLD) & (high.qval >= PLOT.FDR.THRESHOLD), knockout],
          name='insignificant change of virulence',
          marker=list(symbol="circle-open", color='darkgrey'),
          type="scatter", mode="markers") %>%
add_trace(x=results[is.neutral==TRUE, abundance.in],
          y=results[is.neutral==TRUE, log2fc],
          text=results[is.neutral==TRUE, knockout],
          name='neutral reference set',
          marker=list(symbol="circle-open", color='darkgreen'),
          type="scatter", mode="markers") %>%
add_trace(x=results[is.neutral==FALSE & !is.finite(low.qval) & !is.finite(high.qval), abundance.in],
          y=results[is.neutral==FALSE & !is.finite(low.qval) & !is.finite(high.qval), log2fc],
          text=results[is.neutral==FALSE & !is.finite(low.qval) & !is.finite(high.qval), knockout],
          name='non-neutral',
          marker=list(symbol="circle-open", color='black'),
          type="scatter", mode="markers") %>%
add_trace(x=results[low.qval <= PLOT.FDR.THRESHOLD, abundance.in],
          y=results[low.qval <= PLOT.FDR.THRESHOLD, log2fc],
          text=results[low.qval <= PLOT.FDR.THRESHOLD, knockout],
          name='significantly reduced virulence',
          marker=list(symbol="circle-open", color='red'),
          type="scatter", mode="markers") %>%
add_trace(x=results[high.qval <= PLOT.FDR.THRESHOLD, abundance.in],
          y=results[high.qval <= PLOT.FDR.THRESHOLD, log2fc],
          text=results[high.qval <= PLOT.FDR.THRESHOLD, knockout],
          name='significantly increased virulence',
          marker=list(symbol="circle-open", color='blue'),
          type="scatter", mode="markers") %>%
layout(xaxis=list(range=c(0, ceiling(max(log10(results$abundance.in), na.rm=TRUE))), type="log",
                  title="input abundance (loss-corr. #UMIs)"),
       yaxis=list(range=c(floor(min(results$log2fc, na.rm=TRUE)),
                          ceiling(max(results$log2fc, na.rm=TRUE))),
                  title=paste0("virulence log2 fold change")),
       legend=list(x=0, y=-0.15, xanchor="left", yanchor="top"))
```

__________
(The grey area of insignificance virulence changes is only approximate -- it does not take any *false discovery rate* (FDR) or *multiple testing* correction into account, and it is computed from an average *loss*, instead of the per-knockout loss percentage computed by [TRUmiCount](http://www.cibiv.at/software/trumicount). Knockouts are called significant or insignificant based on a FDR threshold of `r PLOT.FDR.THRESHOLD`)

## Reduced virulence compared to the neutral reference set

The following table lists p-values and (FDR-corrected) q-values for the virulence log2 fold-change being significantly smaller than zero, i.e. for the virulence being reduced compared to the neutral reference set

```{r table_depleted}
DT::datatable(results[!is.na(log2fc), list(
    knockout, is.neutral, gene_name, log2fc, pval=low.pval, qval=low.qval, sig=sig.stars(low.qval)
  )],
  filter="top", rownames=FALSE,
  options=list(order=list(3, 'asc'),
               searchCols = list(NULL, NULL, NULL, NULL, NULL, list(search='0 ... 0.1'), NULL))
) %>%
formatRound(columns=c("log2fc"), digits=2) %>%
formatSignif(columns=c("pval", "qval"), digits=2)
```

__________
(*log2fc* contains the log2 fold-changes of the virulence compared to the neutral reference set, *pval* contains p-values for the significance of *log2fc* < 0 under a negative binomial model, *qval* contains FDR-corrected p-values, in *sig* \*\*\*\*\* means *qval* <= 0.00001, \*\*\*\* means *qval* <= 0.0001, \*\*\* means *qval* <= 0.001, \*\* means *qval* <= 0.01, \* means *qval* <= 0.05, + means *qval* <= 0.1. By default only knockouts with *qval* <= 0.1 are shown, remove the column filter to show all)

## Increased virulence compared to the neutral reference set

The following table lists p-values and (FDR-corrected) q-values for the virulence log2 fold-change being significantly larger than zero, i.e. for the virulence being increased compared to the neutral reference set

```{r table_enriched}
DT::datatable(results[!is.na(log2fc), list(
  knockout, is.neutral, gene_name, log2fc, pval=high.pval, qval=high.qval, sig=sig.stars(high.qval)
  )],
  filter="top", rownames=FALSE,
  options=list(order=list(3, 'desc'),
               searchCols = list(NULL, NULL, NULL, NULL, NULL, list(search='0 ... 0.1'), NULL))
) %>%
formatRound(columns=c("log2fc"), digits=2) %>%
formatSignif(columns=c("pval", "qval"), digits=2)
```

__________
(*log2fc* contains the log2 fold-changes of the virulence compared to the neutral reference set, *pval* contains p-values for the significance of *log2fc* > 0 under a negative binomial model, *qval* contains FDR-corrected p-values, in *sig* \*\*\*\*\* means *qval* <= 0.00001, \*\*\*\* means *qval* <= 0.0001, \*\*\* means *qval* <= 0.001, \*\* means *qval* <= 0.01, \* means *qval* <= 0.05, + means *qval* <= 0.1. By default only knockouts with *qval* <= 0.1 are shown, remove the column filter to show all)
